---
title: "Adaptive estimation of spatio-temporal intensities using **kernstadapt**"
output: rmarkdown::pdf_document
geometry: margin = 2cm
fontsize: 11pt
vignette: >
  %\VignetteIndexEntry{spatiotemporal_intenstiy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

```{r, message=FALSE}
# Main package
library(kernstadapt)

# Complementary packages
library(spatstat)
library(sparr)
```


# Introduction

**kernstadapt** is an R package for the spatio-temporal estimation of the intensities of spatio-temporal point processes. The package uses an adaptive approach where the kernel bandwidth varies with each data point to estimate the expected number of points in an observation window and a time interval. **kernstadapt** provides tools for separability testing, adaptive bandwidth computation, and fast calculation of the intensity using a bandwidth partitioning algorithm.


## Intensity function

The intensity function of a spatio-temporal point process gives the expected number of points per unit area in a given spatial region $W$ when the observations depend on time. That is, the intensity functions denotes the expected number of the process events at $\mathbf{u}\in W$ and $v \in T$:

$$
\lambda(\mathbf{u},v)=\lim\limits_{|d \mathbf{u} \times d v| \rightarrow 0}\frac{\mathbb{E} \left[ N(d \mathbf{u}\times d v) \right ]}{d \mathbf{u} d v},
$$
where $d{\mathbf{u}}$ and $d v$ define small spatial ante temporal regions around the locations $\mathbf{u}$ and $v$; where $|\cdot|$ denotes area, and $N(\cdot)$ denotes the number of events of a given set.

 
## Separable estimation of the intensity

A spatio-temporal point process is defined as *first-order separable*, when the intensity function can be factorised as a product of two spatial and temporal parts:

$$
\lambda(\mathbf{u},v) = \lambda_1(\mathbf{u}) \lambda_2(v),
$$
where 
$\lambda_1(\cdot)$ and $\lambda_2(\cdot)$ are non-negative functions.

## Spatio-temporal adaptative estimator

The spatio-temporal adaptive kernel estimator is defined as follows.

$$
		\hat{\lambda}_{\epsilon, \delta}\left(\mathbf{u},v \right) =\frac{1}{e_{\epsilon,\delta} \left(\mathbf{u},v\right) }\sum_{i=1}^{n}{K^{\text{s}}_{\epsilon(\mathbf{u}_i)}\left(\mathbf{u}-\mathbf{u}_{i}\right)  K^{\text{t}}_{\delta(v_i)}\left(v-v_{i}\right) },\qquad (\mathbf{u}, v) \in W\times T,
$$
where the edge correction is given by

$$
e_{\epsilon,\delta} \left(\mathbf{u},v\right)  = \int_{W}\int_{T}K^{\text{s}}_{\epsilon(\mathbf{u'})}\left(\mathbf{u}-\mathbf{u'}\right)  K^{\text{t}}_{\delta(v')}\left(v-v'\right) d \mathbf{u'} d v'.
$$
In above equations, $K^{\text{s}}_{\epsilon(\mathbf{u}_i)}(\cdot)$ and $K^{\text{t}}_{\delta(v_i)}(\cdot)$ are spatial and temporal Gaussian kernels for space and time with bandwidths given by $\epsilon(\mathbf{u}_i)$ and $\delta(v_i) >0$. In this case, since the bandwidths are functions of the data points, these functions must be adequately defined and estimated.

## Fast estimation through partitioning algorithm

The partitioning algorithm performs the computation of the adaptive estimator by summing fixed-bandwidth estimates operating on appropriate subsets of the data points.
The subsets of data points for the partitioning algorithm are defined by binning the variable bandwidths.
The bins generate a partition of the point pattern $X$ into $C_1\times C_2$ subpatterns $Y_{ij}$ and
$X=\bigcup_{ij} Y_{ij}.$ Then, we can compute the intensity as 
$$
\hat{\lambda}_{\epsilon,\delta}\left(\mathbf{u}, v \right) \approx \sum_{i =1}^{C_1} \sum_{j=1}^{C_2} \hat{\lambda}^{*}_{\bar{\epsilon}_i,\bar{\delta}_j}\left(\mathbf{u},v | Y_{ij} \right),
$$
where $\bar{\epsilon}_i$ and $\bar{\delta}_j$ are the midpoints of the $i$th spatial and $j$th temporal bins and $\hat{\lambda}^{*}_{\bar{\epsilon}_i,\bar{\delta}_j}\left(\mathbf{u},v | Y_{ij} \right)$ is a fixed-bandwidth intensity estimate of $Y_{ij}$.



# Data

**kernstadapt** has three datasets included to serve as working examples of the package capabilities.

### Aegiss

This dataset is a spatio-temporal point pattern where the points stand for non-specific gastrointestinal infections in Hampshire, UK. The time covers from 2001 to 2003.

### Santander's earthquakes

This dataset is a spatio-temporal point pattern where the points are earthquakes in Santander, Colombia, from 2000 to 2020.

### Amazon fires

This dataset is a spatio-temporal point pattern where the points are locations of active deforestation fires (starting within past 24 hours) from 01/01/2021 to 10/10/2021 (284 days).


```{r, fig.height = 3, fig.align="center", fig.width=9}
data(aegiss, santander, amazon)
par(mfrow = c(1,3))

plot(aegiss, main = "Aegiss", bg = rainbow(250))
plot(santander, main = "Santander", bg = rainbow(250))
plot(amazon[sample.int(amazon$n, 5000)], main = "Amazon fires", bg = rainbow(250))
```

# Variable bandwidth 

In the adaptive kernel estimation, each data point is equipped with a bandwidth value following the spatial and temporal functions given by

$$
\epsilon(\mathbf{u})=\frac{\epsilon^{\star}}{\gamma^{\text{s}}} \sqrt{\frac{n}{\lambda^{\text{s}}(\mathbf{u})}}, \quad \text{and} \quad
\delta(v)=\frac{\delta^{\star}}{\gamma^{\text{t}}} \sqrt{\frac{n}{\lambda^{\text{t}}(v)}}, $$
where $\epsilon^{\star}$ and $\delta^{\star}$ are *global bandwidths*, $\lambda^{\text{s}}(\mathbf{u})$ and $\lambda^{\text{t}}(v)$ are marginal intensity functions in space and time, $\gamma^{\text{s}}$ and $\gamma^{\text{t}}$ are the geometric means of the marginal intensities.

Here, we assign some global bandwidths using several methods, but the user may let **kernstadapt** choose its defaults.

```{r}
# Cronie and van Lieshout's spatial bandwidth
bw.xy.aegiss <- bw.abram(aegiss, h0 = bw.CvL(santander)) 

# Modified Silverman's rule of thumb temporal bandwidth
bw.t.aegiss <- bw.abram.temp(aegiss$marks, h0 = bw.nrd(aegiss$marks))

# Scott’s isotropic rule of thumb for spatial bandwidth
bw.xy.santander <- bw.abram(santander, h0 = bw.scott.iso(santander))

# Unbiased cross-validation for temporal bandwidth
bw.t.santander <- bw.abram.temp(santander$marks,
                                h0 = bw.ucv(as.numeric(santander$marks)))
```

# Separability test

In order to test separability in a spatio-temporal point process, **kernstadapt** uses a simple statistical test based on spatio-temporal quadrat counts.

```{r}
sapply(list(aegiss, santander, amazon), separability.test)
```

Therefore, we conclude that only Santander dataset can be assumed as separable.


# Separable estimation of the intensity

## Estimator

In the separable case, the estimator is given by
$$
\hat{\lambda}_{\epsilon, \delta}\left(\mathbf{u} ,v\right) =\frac{1}{n}\left( \frac{1}{e_{\epsilon} \left(\mathbf{u}\right) } \sum_{i=1}^{n}{K^{\text{s}}_{\epsilon(\mathbf{u}_i)}\left(\mathbf{u}-\mathbf{u}_{i}\right)}\right) \left(\frac{1}{e_{\delta} \left(v\right)} \sum_{i=1}^{n} K^{\text{t}}_{\delta(v_i)}\left(v-v_{i}\right) \right),\qquad (\mathbf{u}, v) \in W\times T,
$$
where $K^{\text{s}}_{\epsilon}(\cdot)$ and $K^{\text{t}}_{\delta}(\cdot)$ are bivariate and univariate kernels for space and time. $e_{\epsilon}\left(\mathbf{u} \right)$ and $e_{\delta}\left(v \right)$ are edge-correction factors.

**kernstadapt** can estimate the intensity directly by applying the above formula or approximating it using a fast partition method. We apply both methods to Santander's data, the separable one.


## Direct estimator

```{r}
# Direct estimation, separable case
lambda <- dens.direct.sep(X = santander, 
                          dimyx = 128, dimt = 64,
                          bw.xy = bw.xy.santander, 
                          bw.t = bw.t.santander)
```


Now, we plot some snapshots of the spatio-temporal estimated intensity.

```{r, fig.align="center", fig.height = 5, fig.width=9}
# We select some fixed times for visualisation
I <- c(12, 18, 23, 64)

# We subset the lists
SDS <- lapply(lambda[I], function(x) (abs(x)) ^ (1/6))

# Transform to spatial-objects-lists
SDS <- as.solist(SDS)

# We generate the plots
plot(SDS, ncols = 4, equal.ribbon = T, box = F,
     main = 'Direct estimation, separable case')
```


## Partition algorithm estimator

```{r}
# Partition algorithm estimation, separable case
lambda <- dens.par.sep(X = santander, 
                       dimyx = 128, dimt = 64,
                       bw.xy = bw.xy.santander, 
                       bw.t = bw.t.santander,
                       ngroups.xy = 20, ngroups.t = 10)
```

Now, we plot some snapshots of the spatio-temporal estimated intensity.

```{r, fig.align="center", fig.height = 5, fig.width=9}
# We select some fixed times for visualisation
I <- c(12, 18, 23, 64)

# We subset the lists
SPS <- lapply(lambda[I], function(x) (abs(x)) ^ (1/6))

# Transform to spatial-objects-lists
SPS <- as.solist(SPS)

# We generate the plots
plot(SPS, ncols = 4, equal.ribbon = T, box = F, 
     main = 'Partition algorithm estimation, separable case')
```

# Non-separable estimation of the intensity

## Estimator

In the non-separable case, an adaptive estimator for the intensity is

$$
\hat{\lambda}_{\epsilon, \delta}\left(\mathbf{u},v \right) =\frac{1}{e_{\epsilon,\delta} \left(\mathbf{u},v\right)} \sum_{i=1}^{n}{K^{\text{s}}_{\epsilon(\mathbf{u}_i)}\left(\mathbf{u}-\mathbf{u}_{i}\right)  K^{\text{t}}_{\delta(v_i)}\left(v-v_{i}\right) },\qquad (\mathbf{u}, v) \in W\times T,
$$
where the edge correction term is,
$$
e_{\epsilon,\delta} \left(\mathbf{u},v\right)  = \int_{W}\int_{T}K^{\text{s}}_{\epsilon(\mathbf{u'})}\left(\mathbf{u}-\mathbf{u'}\right)  K^{\text{t}}_{\delta(v')}\left(v-v'\right) d \mathbf{u'} d v'.
$$

The **kernstadapt** package has the functionality of applying the direct estimator given above.
It also has a fast partition method for estimating the intensity in a non-separable way.


## Direct estimator

```{r}
#Direct estimation, non-separable case
lambda <- dens.direct(aegiss, 
                      dimyx = 32, dimt = 16,
                      bw.xy = bw.xy.aegiss, 
                      bw.t = bw.t.aegiss,
                      at = "bins")
```

We have used very coarse spatial and temporal grids; as the direct estimator needs lots of computational resources. 

Now, we plot some snapshots of the spatio-temporal estimated intensity.

```{r, fig.align="center", fig.height = 3, fig.width=9}
# We select some fixed times for visualisation
I <- c(2, 5, 8, 16)

# We subset the lists
NSDA <- lapply(lambda[I], function(x) (abs(x)) ^ (1/6))

# Transform to spatial-objects-lists
NSDA <- as.solist(NSDA)

# We generate the plots
plot(NSDA, ncols = 4, equal.ribbon = T, box = F,
     main = 'Direct estimation, non-separable case')
```

## Partition algorithm estimator

We apply the partition algorithm method and let the package decide about global bandwidths.


```{r}
# Partition algorithm estimation, non-separable case
lambda <- dens.par.sep(X = amazon, 
                       dimyx = 128, dimt = 64,
                       ngroups.xy = 20, ngroups.t = 10)
```

Now, the visualisation of the spatio-temporal estimated intensity.

```{r, fig.align="center", fig.height = 3, fig.width=9}
# We select some fixed times for visualisation
I <- c(12, 18, 23, 64)

# We subset the lists
NSPA <- lapply(lambda[I], function(x) (abs(x)) ^ (1/6))

# Transform to spatial-objects-lists
NSPA <- as.solist(NSPA)

# We generate the plots
plot(NSPA, ncols = 4, equal.ribbon = T, box = F,
     main = 'Partition algorithm estimation, non-separable case')
```
