---
title: "Adaptive estimation of spatio-temporal intensities using **kernstadapt**"
output: rmarkdown::pdf_document
geometry: margin = 2cm
vignette: >
  %\VignetteIndexEntry{st_intensity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kernstadapt)
library(spatstat)
library(sparr)
```

# Data
**kernstadapt** has three datasets included to serve as working examples of the package capabilities.

### Aegiss
This dataset is a spatio-temporal point pattern where the points stand for non-specific gastrointestinal infections in Hampshire, UK. The time covers from 2001 to 2003.

### Santander's earthquakes
This dataset is a spatio-temporal point pattern where the points are earthquakes in Santander-Colombia from 2000 to 2020.

### Amazon fires
This dataset is a spatio-temporal point pattern where the points are locations of active deforestation fires (starting within past 24 hours) from 01/01/2021 to 10/10/2021 (284 days).

```{r, fig.height = 3.5, fig.align="center", fig.width=9}
data(aegiss, santander, amazon)
par(mfrow = c(1,3))

plot(aegiss, main = "Aegiss", bg = rainbow(250))
plot(santander, main = "Santander", bg = rainbow(250))
plot(amazon[sample.int(amazon$n, 5000)], main = "Amazon fires", bg = rainbow(250))
```

# Variable bandwidth 

In the adaptive kernel estimation, each data point is equipped with a bandwidth value following the spatial and temporal functions given by
$$
\epsilon(\mathbf{u})=\frac{\epsilon^{\star}}{\gamma^{\text{s}}} \sqrt{\frac{n}{\lambda^{\text{s}}(\mathbf{u})}}, \quad \text{and} \quad
\delta(v)=\frac{\delta^{\star}}{\gamma^{\text{t}}} \sqrt{\frac{n}{\lambda^{\text{t}}(v)}}, $$
where $\epsilon^{\star}$ and $\delta^{\star}$ are *global bandwidths*, $\lambda^{\text{s}}(\mathbf{u})$ and $\lambda^{\text{t}}(v)$ are marginal intensity functions in space and time, $\gamma^{\text{s}}$ and $\gamma^{\text{t}}$ are the geometric means of the marginal intensities.

In this viggnette, we assign some global bandwidths using several methods, but the user may let **kernstadapt** choose its defaults.

```{r}
# Cronie and van Lieshout's spatial bandwidth
bw.xy.aegiss <- bw.abram(aegiss, h0 = bw.CvL(santander)) 
# Modified Silverman's rule of thumb temporal bandwidth
bw.t.aegiss <- bw.abram.temp(aegiss$marks, h0 = bw.nrd(aegiss$marks))

# Scott’s isotropic rule of thumb for spatial bandwidth
bw.xy.santander <- bw.abram(santander, h0 = bw.scott.iso(santander))
# Unbiased cross-validation for temporal bandwidth
bw.t.santander <- bw.abram.temp(santander$marks,
                                h0 = bw.ucv(as.numeric(santander$marks)))
```

# Separable estimation of the intensity

A spatio-temporal point process is defined as *first-order separable*, when the intensity function can be factorised as a product of two spatial and temporal parts:
$$
\lambda(\mathbf{u},v) = \lambda_1(\mathbf{u}) \lambda_2(v),
$$
where 
$\lambda_1(\cdot)$ and $\lambda_2(\cdot)$ are non-negative functions.

## Separability test
In order to test separability in a spatio-temporal point process, **kernstadapt** uses a simple statistical test based on spatio-temporal quadrat counts.

```{r}
sapply(list(aegiss, santander, amazon), separability.test)
```
Theferore, we conclude that only santander dataset can be assumed as separable.

## Estimator
In the separable case, the estimator is given by
$$
\hat{\lambda}_{\epsilon, \delta}\left(\mathbf{u} ,v\right) =\frac{1}{n}\left( \frac{1}{e_{\epsilon} \left(\mathbf{u}\right) } \sum_{i=1}^{n}{K^{\text{s}}_{\epsilon(\mathbf{u}_i)}\left(\mathbf{u}-\mathbf{u}_{i}\right)}\right) \left(\frac{1}{e_{\delta} \left(v\right)} \sum_{i=1}^{n} K^{\text{t}}_{\delta(v_i)}\left(v-v_{i}\right) \right),\qquad (\mathbf{u}, v) \in W\times T,
$$
where $K^{\text{s}}_{\epsilon}(\cdot)$ and $K^{\text{t}}_{\delta}(\cdot)$ are bivariate and univariate kernels for space and time. $e_{\epsilon}\left(\mathbf{u} \right)$ and $e_{\delta}\left(v \right)$ are edge-correction factors.

**kernstadapt** can estimate the intensity directly by applying the above formula or approximating it using a fast partition method. We apply both methods to Santander's data, the separable one.

```{r}
# Direct estimation, separable case
lambda.sep.direct.santander <- dens.direct.sep(X = santander, 
                                               dimyx = 128, dimt = 64,
                                               bw.xy = bw.xy.santander, 
                                               bw.t = bw.t.santander)

# Partition algorithm estimation, separable case
lambda.sep.par.santander <- dens.par.sep(X = santander, 
                                         dimyx = 128, dimt = 64,
                                         bw.xy = bw.xy.santander, 
                                         bw.t = bw.t.santander,
                                         ngroups.xy = 20, ngroups.t = 10)
```

Now, we plot some snapshots of the spatio-temporal estimated intensity.

```{r, fig.align="center", fig.height = 5, fig.width=9}
# We select some fixed times for visualisation
I <- c(12, 18, 23, 64)

# We subset the lists
SDS <- lapply(lambda.sep.direct.santander[I], function(x) (abs(x)) ^ (1/6))
SPS <- lapply(lambda.sep.par.santander[I], function(x) (abs(x)) ^ (1/6))

# Transform to spatial-objects-lists
SDS <- as.solist(SDS)
SPS <- as.solist(SPS)

# We generate the plots
plot(SDS, ncols = 4, equal.ribbon = T, box = F,
     main = 'Direct estimation, separable case')
plot(SPS, ncols = 4, equal.ribbon = T, box = F, 
     main = 'Partition algorithm estimation, separable case')
```


# Non-separable estimation of the intensity

## Estimator
In the non-separable case, an adaptive estimator for the intensity is
$$
\hat{\lambda}_{\epsilon, \delta}\left(\mathbf{u},v \right) =\frac{1}{e_{\epsilon,\delta} \left(\mathbf{u},v\right)} \sum_{i=1}^{n}{K^{\text{s}}_{\epsilon(\mathbf{u}_i)}\left(\mathbf{u}-\mathbf{u}_{i}\right)  K^{\text{t}}_{\delta(v_i)}\left(v-v_{i}\right) },\qquad (\mathbf{u}, v) \in W\times T,
$$
where the edge correction term is,
$$
e_{\epsilon,\delta} \left(\mathbf{u},v\right)  = \int_{W}\int_{T}K^{\text{s}}_{\epsilon(\mathbf{u'})}\left(\mathbf{u}-\mathbf{u'}\right)  K^{\text{t}}_{\delta(v')}\left(v-v'\right) d \mathbf{u'} d v'.
$$

The **kernstadapt** package has the functionality of applying the direct estimator given above.
```{r}
#Direct estimation, non-separable case
lambda.nsep.direct.aegiss <- dens.direct(aegiss, 
                                         dimyx = 32, dimt = 16,
                                         bw.xy = bw.xy.aegiss, 
                                         bw.t = bw.t.aegiss,
                                         at = "bins")
```

We have used very coarse spatial and temporal grids; as the direct estimator needs lots of computational resources. 

Now, we plot some snapshots of the spatio-temporal estimated intensity.

```{r, fig.align="center", fig.height = 3, fig.width=9}
# We select some fixed times for visualisation
I <- c(2, 5, 8, 16)

# We subset the lists
NSDA <- lapply(lambda.nsep.direct.aegiss[I], function(x) (abs(x)) ^ (1/6))

# Transform to spatial-objects-lists
NSDA <- as.solist(NSDA)

# We generate the plots
plot(NSDA, ncols = 4, equal.ribbon = T, box = F,
     main = 'Direct estimation, non-separable case')
```

## Partition algorithm estimator

**kernstadapt** has a fast way for estimating the intensity in a non-separable way, it employs a fast partition method. We apply this method to Amazon data, the most complex dataset of the package. In this case we let the package decide about global bandwidths.

```{r}
# Partition algorithm estimation, non-separable case
lambda.nsep.par.amazon <- dens.par.sep(X = amazon, 
                                       dimyx = 128, dimt = 64,
                                       ngroups.xy = 20, ngroups.t = 10)
```

Now, the visualisation of the spatio-temporal estimated intensity.

```{r, fig.align="center", fig.height = 3, fig.width=9}
# We select some fixed times for visualisation
I <- c(12, 18, 23, 64)

# We subset the lists
NSPA <- lapply(lambda.nsep.par.amazon[I], function(x) (abs(x)) ^ (1/6))

# Transform to spatial-objects-lists
NSPA <- as.solist(NSPA)

# We generate the plots
plot(NSPA, ncols = 4, equal.ribbon = T, box = F,
     main = 'Partition algorithm estimation, non-separable case')
```
